Human Factors Influencing Visual Statistical Inference

### 2013-08-23 sociology_paper.tex
changed p value and discussion of distribution for visual inference

{\bf Keywords: \sf statistical graphics, non-parametric test, cognitive phycology, data visualization, exploratory data analysis, data mining, visual analytics.} 
\end {abstract}

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis.

What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is generated as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true the actual data plot should look just like any of the other plots, and not be identified by an observer. Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A binomial distribution is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  Observers should not be aware of the data that constitutes a lineup, and should not have seen the actual data plot before seeing the lineup. This is the reason that in the election example, above, the scenario was explained after the lineup question, in the text.

@@

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of srtucture, enabling the the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the oberver is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations. (\citet{buja:2009} has a very complete explanation.)

>>>>

{\bf Keywords: \sf statistical graphics, non-parametric test, cognitive \hh{psychology}, data visualization, exploratory data analysis, data mining, visual analytics.} 
\end {abstract}

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are showing from data generated under a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 \hh{out} of 72 observers picked the  plot \hh{of the observed data} (located in spot number $3^2+4$ in the lineup).  The corresponding $p$-value is \hh{0.0077}, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

\hh{How do we interpret this finding, though?} For that, we need to know the context of the data and we need to have more information about the \hh{how null plots were generated}. 
This particular example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is generated as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true the actual data plot should look just like any of the other plots, and not be identified by an observer. Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A distribution \hh{similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario,} is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  Observers should not be aware of the data that constitutes a lineup, and should not have seen the actual data plot before seeing the lineup. This is the reason that in the election example, above, the scenario was explained after the lineup question, in the text.

@@

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations. (\citet{buja:2009} has a very complete explanation.)
